---
title: "Clustering"
output: html_document
---

# Hierarchical clustering (HC)


In this analysis we will see how to make a hierarchical clustering model. The
objectives are to explore if there are similar groups of elements and to see how 
they are interrelated .
 
As example, we will use the Doubs database, based on a sample of fish from the
Doubs river in France. 
![](https://www.google.co.jp/url?sa=i&rct=j&q=&esrc=s&source=images&cd=&cad=rja&uact=8&ved=2ahUKEwinrJSzs4TcAhXRMt4KHdL9BhUQjRx6BAgBEAU&url=https%3A%2F%2Fen.wikipedia.org%2Fwiki%2FDoubs_(river)&psig=AOvVaw03fFZN7MoaB0cPY2agXzJO&ust=1530757814912338)

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ade4)
library(vegan)
library(gclus)
library(cluster)
library(RColorBrewer)
library(labdsv)
```


## 1- Data preparation
We need to remove the rows with double 0 or NA and then normalize the values.

```{r}
data(doubs)
spe <- doubs$fish[-8,]
env <- doubs$env[-8,]
spa <- doubs$xy[-8,]

spe.norm <- decostand(spe, "normalize") 
spe.ch <- vegdist(spe.norm, "euc") 
```


# 2- Choosing the clustering method.
```{r echo=FALSE}
par(mfrow = c(2, 3)) 
```

### Single method

```{r}
spe.ch.single <- hclust(spe.ch, method = "single") 
plot(spe.ch.single, sub = "Single method")
```

### Complete method

```{r}
spe.ch.complete <- hclust(spe.ch, method = "complete")
plot(spe.ch.complete, sub = "Complete method") 
```

### UPGMA method

```{r}
spe.ch.UPGMA <- hclust(spe.ch, method = "average") 
plot(spe.ch.UPGMA, sub = "UPGMA method")
```

### Centroid method

```{r}
spe.ch.centroid <- hclust(spe.ch, method = "centroid") 
plot(spe.ch.centroid, sub = "Centroid method")
```

### Ward method

```{r}
spe.ch.ward <- hclust(spe.ch, method = "ward.D") 
plot(spe.ch.ward, sub = "Ward method")
```

### Squared transformation

```{r}
spe.ch.ward$height <- sqrt(spe.ch.ward$height)
plot(spe.ch.ward, sub = "Squared height method")
```


## 3- Choosing the best method
We can see overfitting in the centroid method case. In the rest of cases we have 
to calculate:
- Cophenetic correlation.
- Gower distance value.

### 3.1. Cophenetic correlation 

```{r}
spe.ch.single.coph <- cophenetic(spe.ch.single) # Single
paste("Single cophenetic:", cor(spe.ch, spe.ch.single.coph))

spe.ch.comp.coph <- cophenetic(spe.ch.complete) # Complete
paste("Complete cophenetic:", cor(spe.ch, spe.ch.comp.coph))

spe.ch.UPGMA.coph <- cophenetic(spe.ch.UPGMA) # Average
paste("UPGMA cophenetic:", cor(spe.ch, spe.ch.UPGMA.coph))

spe.ch.ward.coph <- cophenetic(spe.ch.ward) # Ward 
paste("Ward cophenetic:", cor(spe.ch, spe.ch.ward.coph))

cor(spe.ch, spe.ch.ward.coph, method = "spearman")
```

Graphical representation of the cophenetic correlation (Shepard diagram)

```{r}
par(mfrow = c(2, 2))
plot(spe.ch, spe.ch.single.coph, xlab = "Chord distance", 
     ylab = "Cophenetic distance", asp = 1, xlim = c(0, sqrt(2)), ylim = c(0, sqrt(2)),
     main = c("Single linkage", paste("Cophenetic correlation =",
                                    round(cor(spe.ch, spe.ch.single.coph), 3))))
abline(0, 1);  lines(lowess(spe.ch, spe.ch.single.coph), col = "red")

plot(spe.ch, spe.ch.comp.coph, xlab = "Chord distance", 
     ylab = "Cophenetic distance", asp = 1, xlim = c(0,sqrt(2)), ylim = c(0,sqrt(2)),
     main = c("Complete linkage", paste("Cophenetic correlation =",
                                      round(cor(spe.ch, spe.ch.comp.coph), 3))))
abline(0, 1);  lines(lowess(spe.ch, spe.ch.comp.coph), col = "red")

plot(spe.ch, spe.ch.UPGMA.coph, xlab = "Chord distance", 
     ylab = "Cophenetic distance", asp = 1, xlim = c(0,sqrt(2)), ylim = c(0,sqrt(2)),
     main = c("UPGMA", paste("Cophenetic correlation =",
                           round(cor(spe.ch, spe.ch.UPGMA.coph), 3))))
abline(0, 1);  lines(lowess(spe.ch, spe.ch.UPGMA.coph), col = "red")

plot(spe.ch, spe.ch.ward.coph, xlab = "Chord distance", 
     ylab = "Cophenetic distance", asp = 1, xlim = c(0,sqrt(2)), 
     ylim = c(0,max(spe.ch.ward$height)),
     main = c("Ward clustering", paste("Cophenetic correlation =",
                                     round(cor(spe.ch, spe.ch.ward.coph), 3))))
abline(0, 1);  lines(lowess(spe.ch, spe.ch.ward.coph), col = "red")
```

The best fitted method is the **UPGMA**, with a cophenetic correlation of .861.

### Gower distance (the lower the better)

```{r}
(gow.dist.single <- sum((spe.ch - spe.ch.single.coph) ^ 2))
(gow.dist.comp <- sum((spe.ch - spe.ch.comp.coph) ^ 2))
(gow.dist.UPGMA <- sum((spe.ch - spe.ch.UPGMA.coph) ^ 2))
(gow.dist.ward <- sum((spe.ch - spe.ch.ward.coph) ^ 2))
```

The best method still is UPGMA.


## 3- Choosing the final number of conglomerates.
To this effect we will use three tests:
- a- Fussion level plot
- b- Silhouette plot
- c- Mantel value

### a- Fussion level

```{r}
dev.off()
plot(spe.ch.ward$height, nrow(spe) : 2, type = "S", 
     main = "Fusion levels - Chord - Ward", 
     ylab = "k (number of clusters)", xlab = "h (node height)", col = "grey")
text(spe.ch.ward$height, nrow(spe) : 2, nrow(spe) : 2, col = "red", cex = 0.8)
```

By the length of the fussion line the best number of k groups must be 2 or 4.

### b- Silhouette plot

```{r}
asw <- numeric(nrow(spe))

for(k in 2:(nrow(spe) - 1)){
  sil <- silhouette(cutree(spe.ch.ward, k = k), spe.ch)
  asw[k] <- summary(sil)$avg.width}
k.best <- which.max(asw)

plot(1: nrow(spe), asw, type="h", 
     main = "Silhouette-optimal number of clusters", 
     xlab = "k (number of groups)", ylab = "Average silhouette width")
axis(1, k.best, paste("optimum", k.best, sep = "\n"), col = "red", font = 2,
     col.axis = "red")
points(k.best, max(asw), pch = 16, col = "red", cex = 1.5)
cat("", "Silhouette-optimal number of clusters k =", k.best, "\n", 
    "with an average silhouette width of", max(asw), "\n")
```

### c- Mantel value

```{r}
grpdist <- function(X){
  require(cluster)
  gr <- as.data.frame(as.factor(X))
  distgr <- daisy(gr, "gower")
  distgr}

kt <- data.frame(k = 1:nrow(spe), r = 0)
for(i in 2:(nrow(spe) - 1)){
  gr <- cutree(spe.ch.ward, i)
  distgr <- grpdist(gr)
  mt <- cor(spe.ch, distgr, method = "pearson")
  kt[i, 2] <- mt}
k.best <- which.max(kt$r)

plot(kt$k, kt$r, type = "h", main = "Mantel-optimal number of clusters", 
     xlab = "k (number of groups)", ylab = "Pearson's correlation")
axis(1, k.best, paste("optimum", k.best, sep = "\n"), col = "red", font = 2,
     col.axis = "red")
points(k.best, max(kt$r), pch = 16, col = "red", cex = 1.5)
cat("", "Mantel-optimal number of clusters k =", k.best, "\n", 
    "with a matrix linear correlation of", max(kt$r), "\n")
```

The first and second method tells us that the best number is k = 2. 
The Mantel value indicates that the number must be 4. 
Since the theory suggests the best number is 4, we will validate our model for k = 4.4


## 4- Final validation of the model *(k = 4)*

```{r}
k <- 4 
cutg <- cutree(spe.ch.ward, k = k)
sil <- silhouette(cutg, spe.ch)
rownames(sil) <- row.names(spe)

plot(sil, main = "Silhouette plot", 
     cex.names = 0.8, col = 2:(k + 1), nmax = 100)
```

The groups are coherent, the second group has two unclassified values though. 


## 5- Final plot 
We will use the hcoplot function created by Francois Gillet (2012)

```{r}
hcoplot <- function(tree, diss, k, 
                      title=paste("Reordered dendrogram from", deparse(tree$call), sep="\n"))
{
  require(gclus)
  gr <- cutree(tree, k=k)
  tor <- reorder.hclust(tree, diss)
  plot(tor, hang=-1, xlab=paste(length(gr),"sites"), sub=paste(k,"clusters"), 
       main=title)
  so <- gr[tor$order]
  gro <- numeric(k)
  for (i in 1:k)
  {
    gro[i] <- so[1]
    if (i<k) so <- so[so!=gro[i]]
  }
  rect.hclust(tor, k=k, border=gro+1, cluster=gr)
  legend("topright", paste("Cluster",1:k), pch=22, col=2:(k+1), bty="n")
}

hcoplot(spe.ch.ward, spe.ch, k = 4)
```



# Non-Hierarchical clustering (NHC)

This time we will make a k-mean clustering model.

- 1- We have to normalize the data. This step is already done in the previous analysis.

- 2- Choosing the conglomerate method.

```{r warning=FALSE}
set.seed(1) 
(spe.kmeans <- kmeans(spe.norm, centers = 4, nstart = 100)) 
```

We created our model with 4 groups, the same as the previous HC model.
Here we have the centroids of each cluster, the groups and the variance of each cluster.

## 3- Choosing the number of conglomerates and validation of the model. 
To do this, we will use the next criteria:
 - Calinski & Harabasz value 
 - Simple structure index (SSI)
 - Sum of squared errors (SSE)
 - Silhouette plot 

```{r}
spe.KM.cascade <- cascadeKM(spe.norm, inf.gr = 2, sup.gr = 10, iter = 100, criterion = "ssi")
spe.KM.cascade$results 
head(spe.KM.cascade$partition)
plot(spe.KM.cascade, sortg = TRUE) 
```

The statistics are not determining. By its SSE, the best number is 2 and by its SSI is
3.

```{r}
(spe.kmeans <- kmeans(spe.norm, centers = 3, nstart = 100)) 
dissE <- daisy(spe.norm) 
sk <- silhouette(spe.kmeans$cl, dissE) 
plot(sk)
```

## 4- Final plot and interpretation

```{r}
spebc.ward.g<-cutree(spe.ch.ward,k=4)
table(spe.kmeans$cluster, spebc.ward.g)
```

They only differ in just one case.

```{r}
clusplot(spe.norm, spe.kmeans$cluster, color = TRUE, shade = TRUE, 
         labels = 2, lines = 0)
```

