---
title: "LDA project"
output: html_document
---


# Linear Discriminant Analysis (LDA)

This is a method of **supervised multivariant classification** and it is one of the 
most powerful tool a data analyst have. With this method we can look for the 
significance of the variables and we can make predictions. 
To make this model we need an explanatory matrix X and a response matrix Y.
We will use the iris database as example.

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=FALSE}
library(ade4)
library(vegan)
library(ellipse)
library(mvnormtest)
library(psych)
library(MASS)
library(klaR)
data(iris)
```

```{r tables-mtcars, echo=FALSE}
knitr::kable(iris[1:5, ], caption = 'A caption of the Iris database')
```

## 1- Standarization and exploration

```{r warning=FALSE}
iris.hle <- decostand(as.matrix(iris[1:4]), "hellinger") 
gr <- cutree(hclust(vegdist(iris.hle, "euc"), "ward.D"), 3)
table(gr)
```
We used k = 3 groups, as the theory suggests. The model classified the data in 3 gropus of similar size.



## 2- Assumptions check 
The LDA model is a parametric model, so there are assumptions to check and control.

###2.1. NA values

```{r}
any(is.na(iris))
```
There are not any NA value.

###2.2. Multivariant homogeneity 

```{r warning=FALSE}
iris.pars <- as.matrix(iris[, 1:4])
iris.pars.d <- dist(iris.pars)
(iris.MHV <- betadisper(iris.pars.d, gr))
anova(iris.MHV)
permutest(iris.MHV)
```

We can not accept the homogeneity assumption. We try to transform the data eliminating
outlier values.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
par(mfrow = c(1, 4))
apply(iris[1:4], 2, boxplot)
```
```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=FALSE}
require(fBasics)
```
```{r warning=FALSE}
skewness(iris[1:4])
outliers::outlier(iris[2])
which(iris[2] >= 4.1 | iris[2] < 2.2) 
iris[c(16, 33, 34, 61), 2] <- mean(iris[,2])

iris2 <- cbind(log(iris[1]), iris[2], iris[3], iris[4])
skewness(iris2[1:4])
```
```{r echo=FALSE, warning=FALSE}
apply(iris2[1:4], 2, boxplot)

```
The two first variables are skewed to the right and the third one is skewed to the left.
```{r warning=FALSE}
iris.pars2 <- as.matrix(iris2)
iris.pars.d2 <- dist(iris.pars2)
(iris.MHV2 <- betadisper(iris.pars.d2, gr))
anova(iris.MHV2)
permutest(iris.MHV2)
```
We can not accept the homogeneity of the sample even after transforming the data. In this case
we should make a non-parametric model like a Quadratic Discriminant Analysis (QDA).

###2.3. Normality 
```{r warning=FALSE}
par(mfrow = c(1, ncol(iris.pars2)))
for(j in 1:ncol(iris.pars2)){
  hist(iris.pars2[,j])}

mshapiro.test(t(iris.pars2))
```
We can not accept the normality assumption too, due to the third and forth variables anormality.

###2.4. Multicollinearity 
```{r}
as.dist(cor(iris.pars2))
faraway::vif(iris.pars2)
```
There is one problem of multicollinearity between Petal.Length - Petal.Width. We will continue
our analysis without the Petal.Length variable to improve the output.
```{r}
iris.pars3 <- iris.pars[, -3]
```
###2.5. Linearity
```{r}
psych::pairs.panels(iris[1:4], gap = 0, bg = c("red", "blue", "green")[iris$Species], pch = 21)
```

There are some non-linear relationships. We should try a KDA or K-mDA model instead. We will 
continue with our LDA model in light of example. 



##3- LDA model

```{r warning=FALSE}
iris.pars3.df <- as.data.frame(iris.pars3)
(iris.lda <- lda(gr ~ Sepal.Length + Sepal.Width + Petal.Width, data = iris.pars3.df))
iris.lda$count
```
* The formula of the model will be:
- $LD1 = 0.653 * Sepal.Length - 2.471 * Sepal.Width + 4.975 * Petal.Width$
- $LD2 = -0.659 * Sepal.Length + 2.819 * Sepal.Width + 1.464 * Petal.Width$
The proporton of trace indicates that with just one LD we achive up to a 99% of discimination.

```{r message=FALSE, warning=FALSE}
library(devtools)
# install_github("fawda123/ggord")
library(ggord)
```
#### Plot 1
```{r fig.align='center'}
ggord(iris.lda, iris$Species)
```

####Plot 2
```{r fig.align='center', message=FALSE, warning=FALSE}
partimat(factor(gr) ~ Sepal.Length + Sepal.Width + Petal.Width, data = iris.pars3.df,
         method = "lda", nplots.vert = 1)

Fp <- predict(iris.lda)$x
(iris.class <- predict(iris.lda)$class)
```

#### Plot 3
````{r}
par(mfrow = c(1, 1))
plot(Fp[, 1], Fp[, 2], type = "n")
text(Fp[, 1], Fp[, 2], row.names(iris), col = c(as.numeric(iris.class) + 1))
abline(v = 0, lty = "dotted")
abline(h = 0, lty = "dotted")
for(i in 1:length(levels(as.factor(gr)))){
  cov <- cov(Fp[gr == i, ])
  centre <- apply(Fp[gr == i, ], 2, mean)
  lines(ellipse(cov, centre = centre, level = 0.95))
}
```

There are some problems between the variable 2 and 3 (versicolor and virginica species).



##4- Canonic discrimination evaluation  
###4.1. Canonic value calculation 
```{r}
iris.lda$svd ^ 2 
100 * iris.lda$svd ^ 2/ sum(iris.lda$svd ^ 2) 
```
The first function can explain a 98% of the total variance of the sample.

###4.2. Canonic correlation
```{r}
punt <- predict(iris.lda)$x
summary(lm(punt ~ gr))
```
Our model can explain a 80% of the variance, so we can say our model can discrimate quite well.

##5- Classification accuracy
```{r}
iris.class <- predict(iris.lda)$class
(iris.table <- table(gr, iris.class))
sum(diag(iris.table))/sum(iris.table)
cohen.kappa(iris.table)
cor.test(gr, as.numeric(iris.class), method = "kendall")
```
Our model could manage to classificate correctly the 94% of the data (CCR = .94). A 2% of the 
succesful predictions were due to chance.


##6- Crossed Validation 
```{r}
iris.lda.jac <- lda(gr ~ Sepal.Length + Sepal.Width + Petal.Width, data = iris.pars3.df, CV = TRUE)
summary(iris.lda.jac)

iris.jac.class <- iris.lda.jac$class
iris.jac.table <- table(gr, iris.jac.class)
diag(prop.table(iris.jac.table, 1))
diag(prop.table(iris.table, 1))
```
The crossed model has the same success rate as the previous model. There is no improvement.



##* Prediction example
Let's try to test the prediction capability of our model.
For instance: Sepal.Length = 5, Sepal.Width = 3.2, Petal.Length = 1.2, Petal.Width = 0.1.

```{r}
pred <- c(5, 3.2, 0.1)
pred <- as.data.frame(t(pred))
colnames(pred) <- colnames(iris.pars3)
(pred.result <- predict(iris.lda, newdata = pred))
```
This example is a sample of setosa iris (group 1).


#Quadratic discriminant analysis (QDA)

This could be considered as the non-parametric version of the PCA. It is specially indicated with data that don't follow a normal distribution. 

```{r warnings=FALSE}
iris.qda <- qda(iris[,-5], iris[,5], CV = TRUE)
(tqda <- table(Original = iris$Species, Prediction = iris.qda$class))
diag(prop.table(tqda, 1))
sum(diag(tqda))/sum(tqda)
```
If we use a QDA we improve the discrimination by a 4% (CCR = .98). The non-parametric model is better since we could not accept the homocedasticity and normality assumptions. 
```{r warnings=FALSE, fig.align='center'}
partimat(factor(gr) ~ Sepal.Length + Sepal.Width + Petal.Width, data = iris.pars3.df,
         method = "qda", nplots.vert = 1)
```